---
title: "Import and representation of MERFISH mouse IBD colon data"
author: 
  - name: Daniela Corbetta    
    affiliation: Department of Statistical Sciences, University of Padova
  - name: Ludwig Geistlinger    
    affiliation: Center for Computational Biomedicine, Harvard Medical School
  - name: Tyrone Lee  
    affiliation: Center for Computational Biomedicine, Harvard Medical School
  - name: Jeffrey Moffitt
    affiliation: Department of Microbiology, Harvard Medical School
  - name: Robert Gentleman
    affiliation: Center for Computational Biomedicine, Harvard Medical School
output:
  BiocStyle::html_document:
    self_contained: yes 
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
vignette: >
  % \VignetteIndexEntry{Mouse ibd colon}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL 
)
```

# Installation

To install the package, start R and enter: 

```{r install, eval = FALSE}
if(!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (BiocManager::version() >= "3.20") {
    BiocManager::install("MerfishData")
} else {
    spdl::info(
        "'MerfishData' requires Bioconductor version 3.20 or later, ",
        "installing development version from Github"
    )
    devtools::install_github("ccb-hms/MerfishData")
}
```

# Setup

After the installation, we proceed by loading the package and additional packages
used in the vignette.

```{r lib, message = FALSE}
library(MerfishData)
library(ExperimentHub)
library(ggplot2)
library(terra)
library(scater)
```

# Data

Gut inflammation involves contributions from immune and non-immune cells,
whose interactions are shaped by the spatial organization of the healthy gut and 
its remodeling during inflammation.
The crosstalk between fibroblasts and immune cells is an important axis in this 
process, but our understanding has been challenged by incomplete cell-type 
definition and biogeography.

To address this challenge, 
[Cadinu et al., 2024](https://doi.org/10.1016/j.cell.2024.03.013) 
used multiplexed error-robust
fluorescence in situ hybridization
(MERFISH) to profile the expression of 943 genes in 1.35 million cells imaged
across the onset and recovery from a mouse colitis model. To assign RNAs to 
individual cells, they used [Baysor](https://github.com/kharchenkolab/Baysor).
They identified diverse cell populations, charted their
spatial organization, and revealed their polarization or recruitment in
inflammation.

This vignette demonstrates how to obtain the MERFISH mouse IBD colon dataset 
from [Cadinu et al., 2024](https://doi.org/10.1016/j.cell.2024.03.013) 
from Bioconductor's [ExperimentHub](https://bioconductor.org/packages/ExperimentHub).

The data was obtained from the 
[datadryad data publication](https://doi.org/10.5061/dryad.rjdfn2zh3).

```{r eh}
eh <- ExperimentHub()
query(eh, c("MerfishData", "Cadinu2024"))
```

## Segmented data

It is also possible to obtain the data in a [SpatialExperiment](https://bioconductor.org/packages/SpatialExperiment), which 
integrates experimental data and cell metadata, and provides designated 
accessors for the spatial coordinates. 

```{r data, message = FALSE}
spe <- MouseColonIBD2024()
spe
```

Inspect the data components:

```{r datacomp}
counts(spe)[1:5,1:5]
logcounts(spe)[1:5,1:5]
colData(spe)
head(spatialCoords(spe))
```

The dataset includes cell type labels with three levels of granularity 
(variables `tier1`, `tier2`, `tier3` in the `colData`).

```{r tiers}
# Less specific
table(spe$tier1)
table(spe$tier2)
# More specific
table(spe$tier3)
```

The data have been collected before the insurgence of colitis 
(`sample_type="Healthy"`) and after some time intervals (3 days, 9 days, 21 days).

```{r stype}
table(spe$sample_type)
```

# Visualization

## Reduced dimensions

Replicate Fig. 1C of the [paper](https://doi.org/10.1016/j.cell.2024.03.013), to 
visualize cell types in the UMAP space:

```{r fig1c, message=FALSE}
plotReducedDim(spe, "UMAP", colour_by = "tier1",
               scattermore=TRUE, rasterise = TRUE) +
  scale_color_manual(values = metadata(spe)$colors_tier1) + 
  labs(color='cell type') 
```

## Spatial organization

We can also replicate Fig 1.D, to see the spatial distribution of cell types
in one slide:

```{r fig1d}
# Filter spatial coordinates for the selected slice ID
slice_coords <- spatialCoords(spe)[spe$mouse_id == "082421_D0_m6" &
                                     spe$sample_type=="Healthy" &
                                     spe$technical_repeat_number=="1" &
                                     spe$slice_id=="2", ]
# Rotate coordinates to have them match the rotation in the paper
slice_coords_vec <- vect(slice_coords, type="points")
slice_coords_r <- spin(slice_coords_vec, 180)
slice_c <- as.data.frame(slice_coords_r, geom="XY")
slice_df <- data.frame(x = slice_c[, 1],
                       y = slice_c[, 2],
                       tier1 = spe$tier1[spe$mouse_id == "082421_D0_m6" &
                                     spe$sample_type=="Healthy" &
                                     spe$technical_repeat_number=="1" &
                                     spe$slice_id=="2"])
# Plot
ggplot(data = slice_df, aes(x = x, y = y, color = tier1)) +
  geom_point(shape = 19, size = 0.5) +
  scale_color_manual(values = metadata(spe)$colors_tier1) +
  theme_bw(10) + guides(colour = guide_legend(override.aes = list(size=2)))+ 
  labs(color='cell type') 
```


Lastly, we can compare the spatial organization of cells before the insurgence
of colitis and at the considered timepoints (after 3, 9, and 21 days):

```{r}
# Filter spatial coordinates for the selected slice ID
slice_coords <- spatialCoords(spe)[(spe$mouse_id == "062921_D0_m3a" &
                                     spe$slice_id=="2") |
                                     (spe$mouse_id == "092421_D3_m1" &
                                     spe$slice_id=="2")|
                                     (spe$mouse_id == "062221_D9_m3" &
                                     spe$slice_id=="2") |
                                     (spe$mouse_id == "082421_D21_m1" &
                                     spe$slice_id=="2"), ]


slice_df <- data.frame(x = scale(slice_coords[, 1], scale=FALSE),
                       y = scale(slice_coords[, 2], scale=FALSE),
                       tier1 = spe$tier1[(spe$mouse_id == "062921_D0_m3a" &
                                     spe$slice_id=="2") |
                                     (spe$mouse_id == "092421_D3_m1" &
                                     spe$slice_id=="2")|
                                     (spe$mouse_id == "062221_D9_m3" &
                                     spe$slice_id=="2") |
                                     (spe$mouse_id == "082421_D21_m1" &
                                     spe$slice_id=="2")],
                       day = spe$sample_type[(spe$mouse_id == "062921_D0_m3a" &
                                     spe$slice_id=="2") |
                                    (spe$mouse_id == "092421_D3_m1" &
                                     spe$slice_id=="2")|
                                     (spe$mouse_id == "062221_D9_m3" &
                                     spe$slice_id=="2") |
                                     (spe$mouse_id == "082421_D21_m1" &
                                     spe$slice_id=="2")])

slice_df$day <- factor(slice_df$day, levels = c("Healthy", "DSS3", "DSS9", "DSS21"))
# Plot
p <- ggplot(data = slice_df, aes(x = x, y = y, color = tier1)) +
  geom_point(shape = 19, size = 0.5) +
  scale_color_manual(values = metadata(spe)$colors_tier1) +
  theme_bw(10) + guides(colour = guide_legend(override.aes = list(size=2)))+ 
  labs(color='cell type') 
p <- p + facet_wrap( ~ day, ncol=2, nrow = 2, scales = "free")
p
```

We can also restrict the analysis to a specific macro-group to see the different
distribution of Tier2 cell types. See an example the different composition
of epithelial cells:

```{r}
slice_df$color <- ifelse(slice_df$tier1 == "Epithelial", 
                         as.character(slice_df$tier2), "grey")

slice_df$color <- factor(slice_df$color)

colored_df <- slice_df[slice_df$tier1 == "Epithelial", ]

p <- ggplot() +
  geom_point(data = slice_df, aes(x = x, y = y), color = "grey", 
             shape = 19, size = 0.1) +
  geom_point(data = colored_df, aes(x = x, y = y, color = tier2), 
             shape = 19, size = 0.1) +
  scale_color_manual(values = metadata(spe)$colors_tier2) +
  theme_bw(10) +
  guides(colour = guide_legend(override.aes = list(size = 2))) +
  labs(color = 'cell type') +
  facet_wrap(~ day, ncol = 2, scales = "free")

p
```

```{r}
slice_df$color <- ifelse(slice_df$tier1 == "Immune", 
                         as.character(slice_df$tier2), "grey")

slice_df$color <- factor(slice_df$color)

colored_df <- slice_df[slice_df$tier1 == "Immune", ]

p <- ggplot() +
  geom_point(data = slice_df, aes(x = x, y = y), color = "grey", 
             shape = 19, size = 0.1) +
  geom_point(data = colored_df, aes(x = x, y = y, color = tier2), 
             shape = 19, size = 0.1) +
  scale_color_manual(values = metadata(spe)$colors_tier2) +
  theme_bw(10) +
  guides(colour = guide_legend(override.aes = list(size = 2))) +
  labs(color = 'cell type') +
  facet_wrap(~ day, ncol = 2, scales = "free")

p
```

# Interactive exploration

The MERFISH mouse hypothalamus dataset is part of the
[gallery of publicly available MERFISH datasets](https://moffittlab.connect.hms.harvard.edu/merfish/merfish_homepage.html).

This gallery consists of dedicated 
[iSEE](https://bioconductor.org/packages/iSEE) and 
[Vitessce](http://vitessce.io/) instances, published on 
[Posit Connect](https://posit.co/products/enterprise/connect/), 
that enable the interactive exploration of different segmentations,
the expression of marker genes, and overlay of
cell metadata on a spatial grid or a microscopy image.

# SessionInfo

```{r sessionInfo}
sessionInfo()
```

